\Name: A Non-Parametric, Hierarchical \\ Wavelength-Calibration Method for a Precision Spectrograph

### 2020-10-19 paper/paper.tex
Final(-ish) Referee Response 1

\begin{abstract}\noindent%
\Name\ is a non-parametric, hierarchical framework for precision wavelength-calibration of spectrographs.  It is designed with the needs of extreme-precision radial velocity (\eprv) in mind, which require that instruments be calibrated or stabilized to better than $10^{-4}$ pixels.  Instruments vary along only a few dominant degrees of freedom, especially \eprv\ instruments, which feature highly stabilized optical systems and detectors.  \Name\ takes advantage of this property by using all calibration data to construct a low-dimensional representation of all accessible calibration states for an instrument.  \Name\ also takes advantage of laser-frequency combs or etalons, which generate a dense set of stable calibration points.  This density permits the use of a non-parametric wavelength solution that can adapt to any instrument or detector oddities better than any parametric model, such as a polynomial.  We demonstrate the success of this method with data from the \textsl{EXtreme PREcision Spectrograph} (\expres), which includes a laser frequency comb.  When wavelengths are assigned to laser comb lines using \name, the RMS of the residuals is about five times lower than wavelengths assigned using polynomial fits to individual exposures.  In a typical test case, radial-velocity measurements of HD~34411 showed a reduction in RMS scatter over a 10-month time baseline from $1.22$ to $1.05\, \mps$.
\end{abstract}

% Add official keywords
\keywords{instrumentation: spectrographs -- instrumentation: detectors -- techniques: spectroscopic -- techniques: radial velocities -- methods: data analysis -- methods: statistical}
%\section*{~}\pagebreak
\section{Introduction} 
Precise, radial-velocity (\eprv) programs have been fruitful in finding and characterizing extra-solar planets \citep[e.g.][]{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $50,000-100,000$, which correspond to line widths on the order of $3000\,\mps$.  The state of the art RV precision reached $1\, \mps$ in 2016 \added{\citep{fischer2016}}.  The newest generation of instruments aim  to reach $0.1\,\mps$ precision, the required precision to detect terrestrial worlds.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrographs are well sampled).  Two next-generation spectrographs, \expres\ and \espresso, have been commissioned for more than a year and are demonstrating $<0.1 \mps$ instrumental errors and $\sim 0.2\, \mps$ errors on stars \citep{pepe2013, jurgenson2016, blackman2020, petersburg2020, brewer2020, mascareno2020}.

Traditionally, wavelength solutions are constructed by fitting a polynomial to lines from a calibration source to describe the relationship between wavelength and pixel for each echelle order \citep{butler1996, lovis2007, cersullo2019}.  In this framework, each calibration image is treated independently.  The returned wavelength solutions worked well at the level of 1 $\mps$ precision.

The move towards 0.1 $\mps$ RV precision, necessitates higher-fidelity calibration data and wavelength models.  These models need to account for high-order spatial variations that can arise from small imperfections in the optics of an instrument and non-uniformity in detector pixel sizes/spacing.  There has been significant effort in using an entire set of calibration images to identify incongruous ThAr lines \citep{coffinet2019} or obtain high-resolution Fourier transform spectra of reference cells \citep{wang2020}.  It has also been found that using a segmented polynomial in the dispersion direction, tuned to capture detector defects, better describes the wavelength solution than a single, continuous polynomial \citep{milakovic2020}.

Here, we propose to simplify and improve calibration programs for \eprv\ hardware systems with two practical yet innovative ideas.  The first flows from the fact that calibration sources---which include arc lamps (in some wavelength ranges), etalons, and laser-frequency combs (\lfc s)---illuminate the spectrograph with very stable, very dense sets of lines; almost every location in the spectrograph image plane is surrounded by nearby, useful calibration lines.  This enables use of a calibration methodology that is \emph{non-parametric}, or not defined by a prescribed, analytic function described by a finite number of parameters :  If every point in the spectrograph detector is sufficiently surrounded by nearby calibration lines, the wavelength solution can, for example, be made simply as an interpolation of the calibration data.  The density of lines removes the need to enforce any functional form for the wavelength solution (such as a continuous ninth-order polynomial, for example).  In some ways, this is a generalization of recent work, which has demonstrated the efficacy of constructing a wavelength solution as multiple, segmented polynomials \citep{milakovic2020}.  A non-parametric approach will improve calibration accuracy by not forcing the choice of a parametric form that may bias the calibration, especially when the chosen function is inappropriate (as, for example, polynomials are at detector edges).

The second simple idea follows from the observation that most physical systems have only a few dominant degrees of freedom, meaning most spectrographs vary along only a small number of axes in ``calibration space'', or the (very high-dimensional) space of all possible wavelength solutions.  \replaced{This is particularly true of \eprv\ instruments, which are equipped with stringent environmental stabilizing, likely reducing the number and extent of the axes along which the instrument can vary.}{This is particularly true of \eprv\ instruments, equipped with stringent environmental stabilizing.  The thermomechanical stability of these instruments reduces the variations they experience to something that can be represented by a low-dimensional framework.}  That is, spectrographs, especially stabilized ones, should have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the calibration space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the small, accessible part of calibration space the spectrograph was located for each exposure.

In the context of probabilistic models, this structure is \emph{hierarchical}:  The calibration data are used not just to determine the wavelength solution at one moment, but also to determine the possible \emph{calibration space} of wavelength solutions at all moments.  In statistics, this concept is often described as \emph{de-noising}:  we can improve calibration by recognizing that every calibration exposure contains information about every other calibration exposure.  Thus every exposure can be improved (i.e., de-noised) with information from every other exposure.

The method we propose here---\name---embodies these ideas.
It is a non-parametric, hierarchical, data-driven method to generate a wavelength model.  By being non-parametric, it delivers enormous freedom to the wavelength solution to match or adapt to any instrument or detector oddities.  By being hierarchical, it restricts that freedom tremendously, but it does so appropriately for the empirically determined variations in the stabilized spectrograph.

@@

The \name\ method is designed to take many calibration images, each containing a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate this information into a full wavelength model applicable to all exposures taken with the instrument.  It operates on the two core ideas; The wavelength solution should be allowed flexibility, but it lives in a very low-dimensional calibration space, where the degrees of freedom are set by the limited kinematics of the spectrograph hardware.  \Name\ therefore assumes that the space of possible calibration states for an instrument is low-dimensional while assuming very little about the forms of those states.

\Name\ assumes dense-enough calibration line coverage with well-fit line centers to provide sufficient constraints on an interpolated wavelength solution across an echelle order.  Upstream errors in line center positions may propagate through \name\ wavelength models.  The required line density is dependent on the required precision of the returned wavelength model; larger spacing between lines offer less constraint and are likely to return worse wavelengths.  We revisit and quantify these conditions in Section \ref{sec:others}.

@@

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the \expres\ pipeline sets the function $f(x,y,m;\theta_{n})$ from Equation \ref{eq:wsol} to a 2D, 9\textsuperscript{th}-order polynomial, where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$ \citep{petersburg2020}.
\begin{equation}
\lambda(x,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
Here, the y-dependence is dropped as, in our framing, this dependence is carried by spectral order $m$.  The line position on the detector is therefore uniquely identified by echelle order $m$ and pixel in the dispersion direction, $x$.  The coefficients $c_{nij}$ are interpolated from the time of calibration exposures to time $t_n$ of a science exposure $n$ by a third-order polynomial of time.  This third-order polynomial is evaluated at the time of non-calibration, science exposures to re-construct the coefficients for a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image obtains its $c_{nij}$ coefficients independently.

@@

If the space of all calibration possibilities is in fact $K$-dimensional (where $K$ is a small integer, i.e 2 or 8 or thereabouts), and if the calibration variations are so small that they can be linearized, then the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} should be low-dimensional.  In \name, we transpose the calibration model---making the position $x$ a function of $\lambda$---into the following form
\begin{equation}
x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)
\quad ,
\label{eq:excl_wsol}
\end{equation}
where
$g_0(\lambda,m)$ is the fiducial or mean or standard calibration of the spectrograph,
the $a_{nk}$ are $K$ scalar amplitudes for each exposure $n$,
and the $g_k(\lambda,m)$ are basis functions expressing the ``directions'' in calibration space that the spectrograph can depart from the fiducial calibration.  The resultant $x(\lambda,m,n)$, a list of calibration line positions for a given exposure, can be regarded as the calibration state of the spectrograph for that exposure.  When this calibration structure is used to deliver a wavelength solution, $x(\lambda,m,n)$ can be inverted back into $\lambda(x,m,n)$ to recover wavelengths for each detector position $x$ and echelle order $m$.

The challenge is to learn these basis functions, $g_k(\lambda,m)$, from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern the basis functions.  In this paper, we present one implementation of \expres\ using principal component analysis (PCA) \added{\citep{pearson1901, jolliffe2016}}.  A PCA is justifiable in the limit where there is very high signal-to-nose ratio, as is usually the case with typical calibration images.  There are many alternatives to PCA for this dimensionality reduction; we return to this point in Section \ref{sec:choices} below.

\subsection{Dimensionality Reduction: De-Noising of Calibration Frames} \label{sec:denoising}
\Name\ will use calibration images to determine 1) the space in which an instrument varies and 2) where in the accessible calibration space the spectrograph existed for each exposure.  For each calibration exposure, $n$, \name\ requires a full list of lines, $(\lambda,m)$ that are expected in each calibration exposure.  Each line is uniquely defined by a combination of echelle order, $m$, and ``true'' or theoretical wavelength, $\lambda$.  There are many strategies for identifying calibration line positions and matching them to their assigned wavelengths; this problem is left out of scope for this work.

@@

For each exposure, $n$, every line, $(\lambda,m)$, has an associated fitted (measured) detector position, $x(\lambda,m,n)$, for example x-pixel in an 2D extracted echelle order.  Fitted line centers that are missing from an exposure (e.g. because the fit failed due to noise, the line is not in its usual echelle order, etc.) can be assigned a \code{NaN} (hardware not-a-number) for that echelle order instead.  Let there be $P$ lines per exposure.  \Name\ reads in a $N \times P$ matrix of line positions for each of $P$ lines for each of $N$ exposures.

@@

\begin{algorithm}
\SetAlgoLined
\KwData{line positions $x(\lambda,m,n)$ for each exposure $n$, with wavelengths $\lambda$ and echelle orders $m$}
\KwResult{Basis vectors of the low-dimensional calibration space $g_k(\lambda,m)$, location of exposures in calibration space expressed by amplitudes $a_{n,k}$}
\While{change in missing or outlier line centers $>$ 0.01\%}
{
	$g_0(\lambda,m) = \overline{x(\lambda,m,n)}$\;
	\added{using Singular-Value Decomposition,} find $U, \Sigma, V$ s.t. $U\Sigma V^* = (x(\lambda,m,n)-g_0(\lambda,m))$\;
	let $a_{n,k} = U\cdot \Sigma$ and $g_k(\lambda,m) = V$\;
	$x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)$ for $x(\lambda,m,n)$ = \code{NaN} where $K$ is a a small integer
	}
\caption{Dimensionality Reduction and De-Noising}
\end{algorithm}

\subsection{Interpolating Calibration Position}
 \label{sec:interp_time}
In \name, the amplitude, $a_{n,k}$, of each principal component is interpolated to determine the calibration state of the spectrograph.  For example, the amplitude can be interpolated with respect to time to recreate the calibration state of the spectrograph at different times,\added{$t_{n'}$}.  The choice of what to interpolate against depends on the dominant contribution to variation in the calibration of the instrument.

@@

After experiments, we found that a cubic-spline interpolation that enforces monotonicity, such as a Piecewise Cubic Hermite Interpolating Polynomial (PCHIP) interpolator, works well for interpolating wavelengths onto pixels.  A cubic spline allows for more flexibility than a parameterized function, while the enforced monotonicity allows the wavelength solution, $\lambda(x',m,n')$, to be invertible and prevents spurious deviations that may befall a cubic spline.  Choices in interpolation scheme, $K$, and other tests are further discussed in Section \ref{sec:choice_wvp}.

@@

LFC lines cover echelle orders 84-124, which contain approximately 19200 calibration lines.  Though our results are primarily based on work with LFC data, there will be some discussion of applications to arc lamps below.   ThAr lines cover all 86 extracted orders of \expres\ (echelle orders 75-160), which include approximately 5300 lines.  For both the LFC and ThAr data, lines that appeared in less than 60\% of exposures were not included in the analysis.  Similarly, exposures with more than 60\% of expected lines missing were cut from the analysis.  A list of echelle orders $m$, line wavelengths $\lambda$, and pixel positions $x$, which have been calculated by the EXPRES pipeline \citep{petersburg2020} for every line of every exposure, is read into \name.

@@

@@

Resultant wavelengths are compared on a line-by-line basis.  The residuals of a wavelength solution represent the difference between the wavelength solution evaluated at the line position of a calibration line, and the assigned theoretical wavelength (i.e. from Equation \ref{eq:lfc} for LFC lines and from a line atlas for ThAr lines) for each line in every exposure.  The reported RMS of a wavelength solution is therefore the per-line RMS, i.e.
\begin{equation}
RMS/line \: [m\,s^{-1}] = \sqrt{\sum_{n=1}^N\sum_{p=1}^P\frac{[ \frac{(\lambda_{n,p,pred.} - \lambda_{p,theory})}{\lambda_{p,theory}} \times c ]^2}{N \times P}}
\label{eq:rms}
\end{equation}
where $\lambda_{p,theory}$ is the theoretical wavelength for line $p$, $\lambda_{n,p,pred.}$ is the wavelength predicted by the constructed wavelength solution for line $p$ in exposure $n$, and residuals from all $P$ lines from all $N$ exposures are used, for a total of $N \times P$ lines.  The difference in wavelength is converted to units of $\mps$, a more intuitive metric for \eprv\ work.

@@

\added{The per-line residuals from the denoising test also exhibit smaller spread than interpolation alone.  This suggests that the spectrograph truly is accurately represented by a low-dimensional model, and recreating line positions using this model gives better line position estimates than treating each exposure independently.  The low-dimensional model does not incorporate noise from individual line measurements.  Returning more precise, denoised line positions results in smaller per-line residuals.}

\begin{figure}[t]
\centering
\includegraphics[width=.45\textwidth]{Figures/all_results.pdf}
\caption{Difference in predicted and theoretical wavelengths for the wavelength calibration tests described in Section \ref{sec:tests}.  The per line RMS as defined in Equation \ref{eq:rms} is given in the top-right corner in each method's corresponding color.  \Name\ returns the smallest spread in residuals.}
\label{fig:testHists}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=0.455\textwidth]{Figures/lineResids2D_col.pdf}
\caption{Residuals of a single LFC exposure plotted with respect to detector position (as defined by echelle order and x-pixel) for parametric and non-parametric wavelength calibration methods.  Each line is colored by the difference between the predicted wavelength and the theoretical wavelength for each line, given in units of $\mps$.  \textbf{Top}: residuals to a classic, polynomial fit.  \textbf{Bottom}: residuals to an interpolated wavelength solution.  High-order structure, i.e. vertical stripes and patchiness, is apparent in the residuals to a polynomial wavelength solution, which assumes smoothness.}
\label{fig:resid2d}
\end{figure}

@@

Figure \ref{fig:waveResids} plots the residuals for a subset of order 94 for both a polynomial-based method and a PCHIP-based method guided by either ThAr lines or LFC lines.   The PCHIP model with ThAr lines (orange, dashed curve) returns huge residuals between two widely-separated ThAr lines that extends out of frame.  The classic, polynomial fit exhibits similar residuals in both amplitude and shape regardless of whether the set of ThAr lines or LFC lines are used.  An interpolated wavelength solution using LFC lines (black, solid curve) exhibits the lowest residuals.

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/dm9_rvs.pdf}
\caption{HD 34411 RVs measured with \expres.  All RVs are shown in blue; nightly binned RVs are over-plotted as orange squares.  The RMS of the data set and the RMS of the binned values are given in the top-left corner.  \textbf{Left}: RVs derived using a polynomial-based wavelength solution.  \textbf{Right}: RVs derived using wavelengths from the implementation of \name\ presented in this paper.}
\label{fig:rvs}
\end{figure*}

@@

\subsection{Dimensionality of the Calibration Space, $K$}
\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/pcsLfc9.pdf}
\caption{\textbf{Top}:  The fiducial calibration of the spectrograph, i.e. the mean line positions for each line throughout the epoch of stability.  The following $3 \times 3$ grid of plots show the first nine principal components constructed using LFC lines.  These principal components represent the basis vectors along which the calibration of the spectrograph can deviate from the fiducial calibration.  For each principal component, or basis vector, each calibration line is plotted according to its echelle order and x-pixel and colored by the value of the basis vector for that line.  Principal components beyond the sixth one become steadily more dominated by noise.}
\label{fig:pcLfc}
\end{figure*}

>>>>

\begin{abstract}\noindent%
\Name\ is a non-parametric, hierarchical framework for precision wavelength-calibration of spectrographs.  It is designed with the needs of extreme-precision radial velocity (\eprv) in mind, which require that instruments be calibrated or stabilized to better than $10^{-4}$ pixels.  Instruments vary along only a few dominant degrees of freedom, especially \eprv\ instruments that feature highly stabilized optical systems and detectors.  \Name\ takes advantage of this property by using all calibration data to construct a low-dimensional representation of all accessible calibration states for an instrument.  \Name\ also takes advantage of laser frequency combs or etalons, which generate a dense set of stable calibration points.  This density permits the use of a non-parametric wavelength solution that can adapt to any instrument or detector oddities better than parametric models, such as a polynomial.  We demonstrate the success of this method with data from the \textsl{EXtreme PREcision Spectrograph} (\expres), which uses a laser frequency comb.  When wavelengths are assigned to laser comb lines using \name, the RMS of the residuals is about five times lower than wavelengths assigned using polynomial fits to individual exposures.  Radial-velocity measurements of HD~34411 showed a reduction in RMS scatter over a 10-month time baseline from $1.22$ to $1.05\, \mps$.
\end{abstract}

% Add official keywords
\keywords{instrumentation: spectrographs -- instrumentation: detectors -- techniques: spectroscopic -- techniques: radial velocities -- methods: data analysis -- methods: statistical}
%\section*{~}\pagebreak
\section{Introduction} 
Precise, radial-velocity programs have been fruitful in finding and characterizing extra-solar planets \citep[e.g.][]{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $50,000-100,000$, which correspond to line widths on the order of $3000\,\mps$.  The state of the art RV precision reached $1\, \mps$ in 2016 \added{\citep{fischer2016}}.  The newest generation of instruments aim  to reach $0.1\,\mps$ precision, the required precision to detect terrestrial worlds.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrographs are well sampled).  Two next-generation spectrographs, \expres\ and \espresso, have been commissioned for more than a year and are demonstrating $<0.1 \mps$ instrumental errors and $\sim 0.2\, \mps$ errors on stars \citep{pepe2013, jurgenson2016, blackman2020, petersburg2020, brewer2020, mascareno2020}.

Traditionally, wavelength solutions are constructed by fitting a polynomial to lines from a calibration source in order to describe the relationship between wavelength and pixel for each echelle order \citep{butler1996, lovis2007, cersullo2019}.  In this framework, each calibration image is treated independently.  The returned wavelength solutions worked well at the level of 1 $\mps$ precision.

The move towards 0.1 $\mps$ RV precision, necessitates higher-fidelity calibration data and wavelength models.  These models need to account for high-order spatial variations that can arise from small imperfections in the optics of an instrument and non-uniformity in detector pixel sizes/spacing.  There has been significant effort in using an entire set of calibration images to identify incongruous ThAr lines \citep{coffinet2019} or obtain high-resolution Fourier transform spectra of reference cells \citep{wang2020}.  It has also been found that using multiple polynomials in the dispersion direction, tuned to capture detector defects, better describes the wavelength solution than a single, continuous polynomial \citep{milakovic2020}.

Here, we propose to simplify and improve calibration programs for \eprv\ hardware systems with two practical yet innovative ideas.  The first flows from the fact that calibration sources---which include arc lamps (in some wavelength ranges), etalons, and laser-frequency combs (\lfc s)---illuminate the spectrograph with very stable, very dense sets of lines--almost every location in the spectrograph image plane is surrounded by nearby, useful calibration lines.  This enables use of a calibration methodology that is \emph{non-parametric}, or not defined by a prescribed, analytic function described by a finite number of parameters:  If every point in the spectrograph detector is sufficiently surrounded by nearby calibration lines, the wavelength solution can, for example, be made simply as an interpolation of the calibration data.  The density of lines removes the need to enforce any functional form for the wavelength solution (such as a continuous ninth-order polynomial).  In some ways, this is a generalization of recent work that has demonstrated the efficacy of constructing a wavelength solution as multiple, segmented polynomials \citep{milakovic2020}.  A non-parametric approach will improve calibration accuracy by not forcing the choice of a parametric form that may bias the calibration, especially when the chosen function is inappropriate (as, for example, polynomials are at detector edges).

The second simple idea follows from the observation that most physical systems have only a few dominant degrees of freedom, meaning most spectrographs vary along only a small number of axes in ``calibration space'', or the (very high-dimensional) space of all possible wavelength solutions.  This is particularly true of \eprv\ instruments, which are equipped with stringent environmental stabilizing.  \replaced{ likely reducing the number and extent of the axes along which the instrument can vary.}{The thermomechanical stability of these instruments reduces the variations they experience to something that can be represented by a low-dimensional framework.}  That is, spectrographs, especially stabilized ones, should have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the calibration space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the small, accessible part of calibration space the spectrograph was located for each exposure.

In the context of probabilistic models, this structure is \emph{hierarchical}:  The calibration data are used not just to determine the wavelength solution at one moment, but also to determine the possible \emph{calibration space} of wavelength solutions at all moments.  In statistics, this concept is often described as \emph{de-noising}:  we can improve calibration by recognizing that every calibration exposure contains information about every other calibration exposure.  Thus, every exposure can be improved (i.e., de-noised) with information from every other exposure.

The method we propose here---\name---embodies these ideas.
It is a non-parametric, hierarchical, data-driven method to generate a wavelength model.  By being non-parametric, it delivers enormous freedom to the wavelength solution to match or adapt to any instrument or detector oddities.  By being hierarchical, it restricts that freedom tremendously, but it does so appropriately for the empirically determined variations in a spectrograph.

@@

The \name\ method is designed to take many calibration images, each containing a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate this information into a full wavelength model applicable to all exposures taken with the instrument.  It operates on two core ideas; The wavelength solution should be allowed flexibility, but it lives in a very low-dimensional calibration space where the degrees of freedom are set by the limited kinematics of the spectrograph hardware.  \Name\ therefore assumes that the space of possible calibration states for an instrument is low-dimensional but assumes very little about the forms of those states.

\Name\ also assumes dense-enough calibration line coverage with well-fit line centers to provide sufficient constraints on an interpolated wavelength solution across an echelle order.  Upstream errors in line center positions may propagate through \name\ wavelength models.  The required line density is dependent on the required precision of the returned wavelength model; larger spacing between lines offer less constraint and are likely to return worse wavelengths.  We revisit and quantify these conditions in Section \ref{sec:others}.

@@

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the \expres\ pipeline sets the function $f(x,y,m;\theta_{n})$ from Equation \ref{eq:wsol} to a 2D, 9\textsuperscript{th}-order polynomial, where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$ \citep{petersburg2020}.
\begin{equation}
\lambda(x,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
Here, the y-dependence is dropped as, in our framing, this dependence is carried by spectral order $m$.  The line position on the detector is therefore uniquely identified by echelle order $m$ and pixel in the dispersion direction, $x$.  The coefficients $c_{nij}$ are interpolated from the time of calibration exposures to time $t_n$ of a science exposure $n$ by a third-order polynomial with respect to time.  This third-order polynomial is evaluated at the time of non-calibration, science exposures to re-construct the coefficients for a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image obtains its $c_{nij}$ coefficients independently.

@@

If the space of all calibration possibilities is in fact $K$-dimensional (where $K$ is a small integer, i.e 2 or 8 or thereabouts), and if the calibration variations are so small that they can be linearized, then the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} should be low-dimensional.  In \name, we transpose the calibration model---making the position $x$ a function of $\lambda$---into the following form
\begin{equation}
x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)
\quad ,
\label{eq:excl_wsol}
\end{equation}
where
$g_0(\lambda,m)$ is the fiducial or mean or standard calibration of the spectrograph,
the $a_{nk}$ are $K$ scalar amplitudes for each exposure $n$, and the $g_k(\lambda,m)$ are basis functions expressing the ``directions'' in calibration space that the spectrograph can depart from the fiducial calibration.  The resultant $x(\lambda,m,n)$, a list of calibration line positions for a given exposure, can be regarded as the calibration state of the spectrograph for that exposure.  When this calibration structure is used to deliver a wavelength solution, $x(\lambda,m,n)$ can be inverted back into $\lambda(x,m,n)$ to recover wavelengths for each detector position $x$ and echelle order $m$ (see Figure \ref{fig:cartoon}).

The challenge is to learn these basis functions, $g_k(\lambda,m)$, from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern the basis functions.  In this paper, we present one implementation of \name\ using principal component analysis (PCA) \added{\citep{pearson1901, jolliffe2016}}.  A PCA is justifiable in the limit where exposures have very high signal-to-nose ratio, as is usually the case with typical calibration images.  There are many alternatives to PCA for this dimensionality reduction; we return to this point in Section \ref{sec:choices} below.

\subsection{Dimensionality Reduction: De-Noising of Calibration Frames} \label{sec:denoising}
\Name\ will use calibration images to determine 1) the space in which an instrument varies and 2) where in the accessible calibration space the spectrograph existed for each exposure.  For each calibration exposure, $n$, \name\ requires a full list of lines, $(\lambda,m)$ that are expected to appear in each calibration exposure.  Each line is uniquely defined by a combination of echelle order, $m$, and ``true'' or theoretical wavelength, $\lambda$.  There are many strategies for identifying calibration line positions and matching them to their assigned wavelengths; this problem is left out of scope for this work.

@@

For each exposure, $n$, every line, $(\lambda,m)$, has an associated fitted (measured) detector position, $x(\lambda,m,n)$, for example x-pixel in an 2D extracted echelle order.  Fitted line centers that are missing from an exposure (e.g. because the fit failed due to noise, the line is not in its usual echelle order, etc.) can be assigned a \code{NaN} (hardware not-a-number) for that exposure instead.  Let there be $P$ lines per exposure.  \Name\ reads in a $N \times P$ matrix of line positions for each of $P$ lines for each of $N$ exposures.

@@

\begin{algorithm}
\SetAlgoLined
\KwData{line positions $x(\lambda,m,n)$ for each exposure $n$, with wavelengths $\lambda$ and echelle orders $m$}
\KwResult{Basis vectors of the low-dimensional calibration space $g_k(\lambda,m)$ and location of exposures in calibration space expressed by amplitudes $a_{n,k}$}
\While{change in missing or outlier line centers $>$ 0.01\%}
{
	$g_0(\lambda,m) = \overline{x(\lambda,m,n)}$\;
	\added{using Singular-Value Decomposition,} find $U, \Sigma, V$ s.t. $U\Sigma V^* = (x(\lambda,m,n)-g_0(\lambda,m))$\;
	let $a_{n,k} = U\cdot \Sigma$ and $g_k(\lambda,m) = V$\;
	$x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)$ for $x(\lambda,m,n)$ = \code{NaN} where $K$ is a a small integer
	}
\caption{Dimensionality Reduction and De-Noising}
\end{algorithm}

\subsection{Interpolating Calibration Position}
 \label{sec:interp_time}
In \name, the amplitude, $a_{n,k}$, of each principal component is interpolated to determine the calibration state of the spectrograph.  For example, the amplitude can be interpolated with respect to time to recreate the calibration state of the spectrograph at different times.  The choice of what to interpolate against depends on the dominant contribution to variation in the calibration of the instrument.

@@

After experiments, we found that a cubic-spline interpolation that enforces monotonicity, such as a Piecewise Cubic Hermite Interpolating Polynomial (PCHIP) interpolator, works well for interpolating wavelengths onto pixels.  A cubic spline allows for more flexibility than a parameterized function, while the enforced monotonicity allows the wavelength solution, $\lambda(x',m,n')$, to be invertible and prevents spurious deviations that may befall a cubic spline.  Choices in interpolation scheme, $K$, and other tests are further discussed in Section \ref{sec:choice_wvp}.  \added{Our implementation of \name\ is hosted on GitHub\footnote{\url{https://www.github.com/davidwhogg/EPRVCalibration}}.}

@@

LFC lines cover echelle orders 84-124, which contain approximately 19200 calibration lines.  Though our results are primarily based on work with LFC data, there will be some discussion of applications to arc lamps below.   ThAr lines cover all 86 extracted orders of \expres\ (echelle orders 75-160), which include approximately 5300 lines.  For both the LFC and ThAr data, lines that appeared in less than 60\% of exposures were not included in the analysis.  Similarly, exposures with more than 60\% of expected lines missing were cut from the analysis.  A list of echelle orders $m$, line wavelengths $\lambda$, and pixel positions $x$ were calculated by the EXPRES pipeline \citep{petersburg2020} for every line of every exposure and read into \name.

@@

\begin{figure}[t]
\centering
\includegraphics[width=.45\textwidth]{Figures/all_results.pdf}
\caption{Difference in predicted and theoretical wavelengths for the wavelength calibration tests described in Section \ref{sec:tests}.  The per line RMS as defined in Equation \ref{eq:rms} is given in the top-right corner in each method's corresponding color.  Incorporating denoising returns the smallest spread in residuals.}
\label{fig:testHists}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=0.455\textwidth]{Figures/lineResids2D_col.pdf}
\caption{Residuals of a single LFC exposure plotted with respect to detector position (as defined by echelle order and x-pixel) for parametric (top) and non-parametric (bottom) wavelength calibration methods.  Each line is colored by the difference between the predicted wavelength and the theoretical wavelength for each line, given in units of $\mps$.  High-order structure, i.e. vertical stripes and patchiness, is apparent in the residuals to a polynomial wavelength solution, which assumes smoothness.}
\label{fig:resid2d}
\end{figure}

@@

The residuals of a wavelength solution represent the difference between the wavelength solution evaluated at the line position of a calibration line, and the assigned theoretical wavelength (i.e. from Equation \ref{eq:lfc} for LFC lines) on a line-by-line basis in every exposure.  The reported RMS of a wavelength solution is therefore the per-line RMS, i.e.
\begin{equation}
RMS/line \: [m\,s^{-1}] = \sqrt{\sum_{n=1}^N\sum_{p=1}^P\frac{[ \frac{(\lambda_{n,p,pred.} - \lambda_{p,theory})}{\lambda_{p,theory}} \times c ]^2}{N \times P}}
\label{eq:rms}
\end{equation}
where $\lambda_{p,theory}$ is the theoretical wavelength for line $p$, $\lambda_{n,p,pred.}$ is the wavelength predicted by the constructed wavelength solution for line $p$ in exposure $n$, and residuals from all $P$ lines from all $N$ exposures are used, for a total of $N \times P$ lines.  The difference in wavelength is converted to units of $\mps$, a more intuitive metric for \eprv\ work.

@@

\added{The per-line residuals from the denoising test also exhibit smaller spread than interpolation alone.  This suggests that the spectrograph truly is accurately represented by a low-dimensional model.  Recreating line positions using this model gives better line position estimates than treating each exposure independently.  The low-dimensional model does not incorporate noise from individual line measurements.  Returning more precise, denoised line positions results in smaller per-line residuals.}

@@

Figure \ref{fig:waveResids} (right) plots the residuals for a subset of order 94 for both a polynomial-based method and a PCHIP-based method guided by either ThAr lines or LFC lines.   The PCHIP model with ThAr lines (orange, dashed curve) returns huge residuals between two widely-separated ThAr lines that extends out of frame.  The classic, polynomial fit exhibits similar residuals in both amplitude and shape regardless of whether the set of ThAr lines or LFC lines are used.  An interpolated wavelength solution using LFC lines (black, solid curve) exhibits the lowest residuals.

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/dm9_rvs.pdf}
\caption{HD 34411 RVs measured with \expres.  We zoom in on all but 8 of the exposures, which were taken over 150 days later.  All RVs are shown in blue; nightly binned RVs are over-plotted as orange squares.  The RMS of the data set and the RMS of the binned values are given in the top-left corner.  \textbf{Left}: RVs derived using a polynomial-based wavelength solution.  \textbf{Right}: RVs derived using wavelengths from the implementation of \name\ presented in this paper.}
\label{fig:rvs}
\end{figure*}

@@

\subsection{Dimensionality of the Calibration Space, $K$}
\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/pcsLfc9.pdf}
\caption{\textbf{Top}:  The fiducial calibration of the spectrograph, i.e. the mean line positions for each line throughout the epoch of stability.  \textbf{The following $3 \times 3$ grid of plots} show the first nine principal components constructed using LFC lines.  These principal components represent the basis vectors along which the calibration of the spectrograph can deviate from the fiducial calibration.  For each principal component, or basis vector, each calibration line is plotted according to its echelle order and x-pixel and colored by the value of the basis vector for that line.  Principal components beyond the sixth one become steadily more dominated by noise.}
\label{fig:pcLfc}
\end{figure*}
